Tropical Mathematics for the Lambda Calculus

### 2023-01-15 section1.tex
commit

In the recent years, more and more interest in the programming language community has been directed towards the analysis of \emph{quantitative} properties of algorithms, such as e.g.~number of computation steps, convergence probability, 
approximation bounds, 
as opposed to purely \emph{qualitative} properties like e.g.~termination or program equivalence.

Notably, a significative effort has been made in order to extend, or adapt, well-established qualitative methods, like type systems, relational logics or denotational semantics, to account for quantitative properties. We can mention, for example, 
intersection type systems aimed at capturing time or space resources \cite{Beniamino}, or convergence probabilities \cite{UgoBreuvart, LICS2022},  relational logics to account for probabilistic properties like e.g.~differential privacy \cite{Barthes} or metric preservation \cite{Reed2010, dallago}, as well as the study of denotational models for 
probabilistic \cite{PCOH, QBS} or differential \cite{difflambda} extensions of the $\lambda$-calculus. 
Usually, the main reason to resort to such logical, type-theoretical or denotational approaches is to obtain methods which are \emph{modular} and \emph{compositional}, that is, where properties of complex algorithms can be deduced from properties of their component parts.

@@

Among such quantitative approaches, two different directions have received a considerable attention in recent years.

On the one hand there is the approach based on \emph{non-idempotent} intersection types \cite{} and their semantical counterpart, the so-called \emph{relational semantics} \cite{}, which have been exploited to account for higher-order program differentiation \cite{difflambda}, to establish reasonable \emph{cost-models} for the $\lambda$-calculus \cite{}, and have also been shown suitable to capture probabilistic properties \cite{Manzo}.

On the other hand there is the approach of \emph{program metrics} \cite{} and \emph{quantitative equational theories} \cite{Mardare}: when considering probabilistic or approximated computations, rather than knowing whether two programs compute \emph{the same} function, which is rarely the case, it is often more useful to know whether they compute functions that do not differ \emph{too much}. This has motivated the study of denotational semantics in which types are endowed with a metric, measuring similarity of behavior; this approach has found  applications in e.g.~differential privacy \cite{} and coinductive methods \cite{}, and was recently extended to account for the the full $\lambda$-calculus \cite{LICS2021, FSCD2022}.

In both approaches the notion of \emph{linearity}, in the sense of linear logic \cite{linearlogic} (i.e.~of using its input exactly once), plays a crucial role.
In the relational semantics linear programs correspond to linear maps, in the usual algebraic sense, while the possibility of duplicating inputs leads to consider \emph{analytic} maps, i.e.~formal infinite series of monomials of the form $a_{n}\cdot x^{n}$.
By contrast, in metric semantics linear programs correspond to \emph{non-expansive} maps, i.e.~to functions that do not increase distances, while the possibility of duplicating inputs leads to interpret \emph{bounded} programs (i.e.~programs with a fixed duplication bound) as \emph{Lipschitz-continuous} maps \cite{Gabo2017}, and general programs as maps satisfying coarser 
continuity conditions (see e.g.~\cite{LICS2021}).

A natural question is thus whether these two, apparently unrelated, ways of interpreting linearity and duplication can be somehow connected. At a first glance, there seems to be a fundamental ``logarithmic'' gap between the two approaches.
In the relational semantics a typical ``duplicating'' map is obtained by composing the diagonal with multiplication:
$$
\begin{tikzcd}
\mathbb R \ar{rrr}{x\mapsto \langle x, x\rangle}
 & &  &
 \mathbb R\times \mathbb R 
 \ar{rrr}{\langle x,y\rangle \mapsto x\cdot y}
 & & & \mathbb R
\end{tikzcd}
$$
yielding the square product function $\lambda x.x^{2}$.
However, in metric semantics this function needs not even exist (as these models are often restricted to Lipschitz-continuous maps \cite{Gabo2017})! Instead, a typical ``duplicating'' map can be obtained by composing the diagonal with the sum 
$$
\begin{tikzcd}
\mathbb R \ar{rrr}{x\mapsto \langle x, x\rangle}
 & &  &
 \mathbb R\times \mathbb R 
 \ar{rrr}{\langle x,y\rangle \mapsto x+y}
 & & & \mathbb R
\end{tikzcd}
$$
yielding the linear (and Lipschitz) function $\lambda x.2x$.

As this example seems to suggest, there seems to be a sort of ``logarithmic'' gap between the two approaches. Can this be made explicit?

\subsection{Tropical Geometry: a ``logarithmic'' version of usual geometry}

In very simple words, tropical geometry describes what happens to usual algebra and geometry when we replace ``$\cdot$'' by  ``$+$'', and ``$+$'' by ``$\min$''. It can thus be seen as a sort of ``logarithmic'' version of usual geometry (this idea can be made precise via the so-called \emph{Maslov deformation} \cite{}).
Tropical geometry is a kind of \emph{idempotent} mathematics, since the role of addition is 
played by the idempotent operation $\min$.
For instance, a quadratic polynomial like e.g.~$p(x)=ax^{2}+bx+c$ corresponds in the tropical word correspond to the \emph{linear} function
$$
\trop p(\alpha)=\min\{2\alpha + a, \alpha+b, c\}
$$

The tropicalization of usual algebraic concepts comes with the important advantage that many intractable properties may become efficiently tractable, in a computational way (e.g.~finding roots of polynomials). For this reason, methods from tropical geometry are more and more applied in computer science, e.g.~for convex analysis and machine learning \cite{}.

Coming back to our previous discussion of program semantics, tropical geometry might seem to provide precisely what we are looking for: a way to see polynomials as linear (i.e.~Lipschitz) functions.
At this point, it is worth mentioning that a tropical variant of relational semantics has already been considered \cite{}, and shown capable of capturing \emph{optimal} probabilistic properties, but has not yet been studied in detail. Furthermore, connections between tropical linear algebra and the abstract study of metric spaces and non-expansive functions, and more generally, of the study of \emph{quantale-enriched} categories, have also been observed \cite{}.

The goal of this paper is to make such intuitive connections precise, yielding a way to relate, conceptually, two different quantitative approaches to higher-order algorithms, as well as suggesting new potential applications of tropical geometry to the study of quantitative properties of programs.

>>>>

In recent years, more and more interest in the programming language community has been directed towards the analysis of \emph{quantitative} properties of algorithms, such as e.g.~number of computation steps, convergence probability, 
approximation bounds, 
as opposed to purely \emph{qualitative} properties like e.g.~termination or program equivalence. 
Notably, a significative effort has been made to extend, or adapt, well-established qualitative methods, like e.g.~type systems, relational logics or denotational semantics, to account for quantitative properties. We can mention, for example, 
intersection type systems aimed at capturing time or space resources \cite{Beniamino}, or convergence probabilities \cite{UgoBreuvart, LICS2022},  relational logics to account for probabilistic properties like e.g.~differential privacy \cite{Barthes} or metric preservation \cite{Reed2010, dallago}, as well as the study of denotational models for 
probabilistic \cite{PCOH, QBS} or differential \cite{difflambda} extensions of the $\lambda$-calculus. 
The main reason to look for methods relying on (quantitative extensions of) type-theory or denotational semantics is that these approaches yield \emph{modular} and \emph{compositional} techniques, that is, allow one to deduce properties of complex algorithms from the properties of their component parts.

@@

Among such quantitative approaches, two different directions have received a considerable attention.

On the one hand there is the approach of \emph{program metrics} \cite{} and \emph{quantitative equational theories} \cite{Mardare}: when considering probabilistic or approximated computations, rather than knowing whether two programs compute \emph{the same} function, which is rarely the case, it is often more useful to know whether they compute functions that do not differ \emph{too much}. This has motivated the study of denotational semantics in which types are endowed with a metric, measuring similarity of behavior; this approach has found  applications in e.g.~differential privacy \cite{} and coinductive methods \cite{}, and was recently extended to account for the the full $\lambda$-calculus \cite{LICS2021, FSCD2022}.

On the other hand there is the approach based on \emph{differential} or \emph{resource-aware} extensions of the $\lambda$-calculus, which is well-connected with the \emph{relational semantics} \cite{} and has a syntactic counterpart in the study of \emph{non-idempotent} intersection types \cite{}. This family of approaches have been exploited to account for higher-order program differentiation \cite{difflambda}, to establish reasonable \emph{cost-models} for the $\lambda$-calculus \cite{}, and have also been shown suitable to capture probabilistic properties \cite{Manzo}.

In both approaches the notion of \emph{linearity}, in the sense of linear logic \cite{linearlogic} (i.e.~of using its input exactly once), plays a crucial role.
In the metric semantics linear programs correspond to \emph{non-expansive} maps, i.e.~to functions that do not increase distances; moreover, the possibility of duplicating inputs leads to interpret \emph{bounded} programs (i.e.~programs with a fixed duplication bound) as \emph{Lipschitz-continuous} maps \cite{Gabo2017}.
By contrast, in the standard semantics of the differential $\lambda$-calculus, linear programs correspond to linear maps, in the usual algebraic sense, while the possibility of duplicating inputs leads to consider functions defined as \emph{power series}.

A natural question is thus whether these two, apparently unrelated, ways of interpreting linearity and duplication can be somehow connected. At a first glance, there seems to be a fundamental ``logarithmic'' gap between the two approaches:
in metric models $n$ times duplication results in a linear (hence Lipschitz) function $n\cdot x$, while in differential models this results to
a monomial $x^{n}$, which is not Lipschitz. The fundamental insight of this article is that a natural way to overcome 
this obstacle 
is to work in the framework of 
tropical mathematics.
%
%''
%
%s
%emantics a typical ``duplicating'' map is obtained by composing the diagonal with multiplication:
%$$
%\begin{tikzcd}
%\mathbb R \ar{rrr}{x\mapsto \langle x, x\rangle}
% & &  &
% \mathbb R\times \mathbb R 
% \ar{rrr}{\langle x,y\rangle \mapsto x\cdot y}
% & & & \mathbb R
%\end{tikzcd}
%$$
%yielding the square product function $\lambda x.x^{2}$.
%However, in metric semantics this function needs not even exist (as these models are often restricted to Lipschitz-continuous maps \cite{Gabo2017})! Instead, a typical ``duplicating'' map can be obtained by composing the diagonal with the sum 
%$$
%\begin{tikzcd}
%\mathbb R \ar{rrr}{x\mapsto \langle x, x\rangle}
% & &  &
% \mathbb R\times \mathbb R 
% \ar{rrr}{\langle x,y\rangle \mapsto x+y}
% & & & \mathbb R
%\end{tikzcd}
%$$
%yielding the linear (and Lipschitz) function $\lambda x.2x$.
%
%As this example seems to suggest, there seems to be a sort of ``logarithmic'' gap between the two approaches. Can this be made explicit?

\subsection{Tropical mathematics and program semantics }

Tropical mathematics was introduced in the seventies by the Brazilian mathematician Imre Simon (whose origin is, as far as we know, the only reason for the ``tropical'' adjective) as a way to do algebra and geometry by replacing the usual ring structure of numbers based on addition and multiplication by the semi-ring structure given by ``$\min$'' and ``$+$''.
%
%
% interpreting the usual ``$\times$'' and  ``$+$'' operations by  ``$+$'' by ``$\min$''. It can thus be seen as a sort of ``logarithmic'' version of usual geometry (this idea can be made precise via the so-called \emph{Maslov deformation} \cite{}).
Tropical mathematics is a form of \emph{idempotent} mathematics, since the role of addition is 
played by the idempotent operation $\min$.
For instance, a quadratic polynomial like e.g.~$p(x)=ax^{2}+bx+c$, when interpreted over the tropical semi-ring, becomes the piecewise linear function
$
\varphi(\alpha)=\min\{2\alpha + a, \alpha+b, c\}
$.

The tropicalization of usual algebraic concepts comes with the important advantage that many intractable properties may become efficiently tractable, in a computational way (e.g.~finding roots of polynomials, or studying geometric invariants like e.g.~the genus). For this reason, methods from tropical geometry are more and more applied in computer science, e.g.~for convex analysis and machine learning \cite{}.

Coming back to our previous discussion of program semantics, tropical geometry might seem to provide precisely what we are looking for, since polynomials become Lipschitz functions.
At this point, it is worth mentioning that a tropical variant of relational semantics has already been considered \cite{}, and shown capable of capturing \emph{optimal} probabilistic properties, but has not yet been studied in detail. Furthermore, connections between tropical linear algebra and the abstract study of metric spaces and non-expansive functions, and more generally, of the study of \emph{quantale-enriched} categories, have also been observed \cite{}.

The goal of this paper is to study the tropical semantics of the differential $\lambda$-calculus and use it to connect it with metric semantics. This bridge does not only provide a conceptual unification of two different quantitative approaches to higher-order algorithms, but it also 
suggests ways in which techniques for resource-analysis could be used for sensitivity analysis and vice-versa, opening the way for new potential applications of tropical geometry to the study of quantitative properties of higher-order programs.
